---
title: "Compositional data analysis with dsCodaClient"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compositional data analysis with dsCodaClient}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The whole vignette runs against DSLite, so it only evaluates when the full
# DataSHIELD stack is installed. Elsewhere (e.g. R CMD check without dsBase)
# the code is shown but not run.
stack <- c("DSLite", "dsBase", "dsBaseClient", "dsCoda", "dsTidyverseClient",
           "compositions", "zCompositions")
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = all(vapply(stack, requireNamespace, logical(1), quietly = TRUE))
)
```

## What this package does

Compositional data (CoDa) are vectors of parts that carry relative
information only: the parts sum to a constant, so increasing one part
necessarily decreases the others, and standard statistics on the raw parts
are misleading. The usual remedy is to work with log-ratios of the parts,
which live in ordinary Euclidean space.

The running example is the 24-hour movement behaviour composition used in
the ProPASS consortium: minutes per day of vigorous physical activity
(`vpa`), moderate physical activity (`mpa`), light physical activity
(`lipa`), sleep (`sleep`) and sedentary behaviour (`sb`), which sum to 1440.

`dsCodaClient` lets you run that workflow across DataSHIELD servers
without seeing individual records. Every server needs `dsBase`, `dsCoda`
and (for the regression step) `dsTidyverse` installed.

## Data requirements

* Parts are columns of a data frame; one row per participant.
* Parts must be strictly positive. Zeros are not allowed in a composition,
  so replace them first (see below).
* Every server must hold the same parts under the same column names.

## A DSLite session with three studies

We simulate three studies. A handful of participants have zero minutes of
vigorous activity, which is what happens in real accelerometer data.

```{r data, message = FALSE}
library(DSI)
library(DSLite)
library(dsBaseClient)
library(dsTidyverseClient)
library(dsCodaClient)

set.seed(1)
make_study <- function(n, zeros) {
  x <- data.frame(
    id = seq_len(n),
    vpa = rgamma(n, 2, 1) * 10 + 1,
    mpa = rgamma(n, 3, 1) * 30 + 1,
    lipa = rgamma(n, 4, 1) * 60 + 1,
    sleep = rnorm(n, 480, 40),
    sb = rnorm(n, 600, 60),
    bmi = rnorm(n, 26, 4)
  )
  x$vpa[sample(n, zeros)] <- 0
  x
}
study1 <- make_study(120, 8)
study2 <- make_study(150, 10)
study3 <- make_study(90, 6)

dslite.server <- newDSLiteServer(
  tables = list(study1 = study1, study2 = study2, study3 = study3),
  config = DSLite::defaultDSConfiguration(include = c("dsBase", "dsCoda", "dsTidyverse"))
)

builder <- DSI::newDSLoginBuilder()
for (s in c("study1", "study2", "study3")) {
  builder$append(server = s, url = "dslite.server", table = s, driver = "DSLiteDriver")
}
conns <- DSI::datashield.login(builder$build(), assign = TRUE, symbol = "D")
```

Against real servers you would build `logindata` with Opal or Armadillo
credentials; everything after `datashield.login()` is identical.

## 1. Replace zeros before closing the composition

`ds.lrEM()` runs the log-ratio EM algorithm of `zCompositions::lrEM()`
on the server. It replaces values flagged by `label` with imputed values
below the detection limits in `dl`, given on the same scale as the data
(minutes here). Do this on the raw parts, before `ds.acomp()`.

```{r zeros, results = "hide"}
parts <- c("vpa", "mpa", "lipa", "sleep", "sb")

ds.select(df.name = "D", tidy_expr = list(vpa, mpa, lipa, sleep, sb),
          newobj = "D_parts", datasources = conns)

ds.lrEM(X = "D_parts", label = 0, dl = rep(1, length(parts)),
        objectname = "D_imp", datasources = conns)
```

```{r zeros-check}
ds.dim("D_imp", datasources = conns)
```

Note that `z.delete = TRUE` (the default) drops rows whose parts are all
zero, so check the dimensions still match `D` before combining objects.

## 2. Close the composition

`ds.acomp()` creates a `compositions::acomp` object on each server. Parts
can be given by name or index; `total = 1` rescales each row to
proportions.

```{r acomp, results = "hide"}
ds.acomp(X = "D_imp", parts = parts, objectname = "comp", datasources = conns)
```

```{r acomp-check}
ds.class("comp", datasources = conns)
```

## 3. Isometric log-ratio coordinates

An ILR transform maps a D-part composition to D - 1 orthonormal
coordinates. `build_sequential_ilr_V()` builds the basis for a
sequential binary partition: coordinate 1 contrasts `vpa` against all
other parts, coordinate 2 contrasts `mpa` against the parts after it, and
so on. Reorder `parts` if you want a different first contrast.

```{r basis}
V <- build_sequential_ilr_V(length(parts), parts)
round(V, 3)
```

`ds.ilr()` sends the basis to the servers and stores the coordinates. With
`V = NULL` the default basis of `compositions::ilrBase()` is used instead.

```{r ilr, results = "hide"}
ds.ilr(X = "comp", V = V, objectname = "ilr_sbp", datasources = conns)
```

## 4. Pooled mean composition

`ds.meanComp()` computes the mean of each ILR coordinate across all
servers with `dsBaseClient::ds.mean(type = "combine")` and inverts the
transform on the client. The result is the pooled geometric-mean
composition, which is the compositional analogue of the arithmetic mean.

```{r mean}
mean_comp <- ds.meanComp("comp", datasources = conns)
round(mean_comp, 4)
round(mean_comp * 1440)  # back to minutes per day
```

## 5. Regression on the ILR coordinates

The coordinates are stored as a `compositions::rmult` matrix, which
`ds.glm()` cannot address by column. Convert it to a tibble first; the
columns come out named `V1` to `V4`, where `V1` is the first column of
`V` (`ilr1`), `V2` is `ilr2`, and so on.

```{r regression, message = FALSE, warning = FALSE}
ds.as_tibble("ilr_sbp", newobj = "ilr_tbl", datasources = conns)
ds.colnames("ilr_tbl", datasources = conns)$study1

fit <- ds.glm(
  formula = "D$bmi ~ ilr_tbl$V1 + ilr_tbl$V2 + ilr_tbl$V3 + ilr_tbl$V4",
  family = "gaussian", datasources = conns
)
round(fit$coefficients, 3)
```

Server-side messages and `glm.fit` convergence warnings from the
single-step IRLS iterations are hidden in this chunk; they are normal for
the DataSHIELD GLM algorithm and stay on the server with a real
connection. The coefficient for `ilr_tbl$V1` is the change in BMI per
unit increase of the first coordinate, i.e. more `vpa` relative to all
other behaviours.

```{r logout}
DSI::datashield.logout(conns)
```
